### 2018-09-06 paper/finitecomplete-failures.tex
Added theorem showing that maximal trace length pq is really ALWAYS required, for all p, q >= 2.

The following example, however, shows that the maximal length $pq$ is really required 
when testing for refinement.
\begin{example}\label{ex:pq}
Consider the CSP reference process $P$ and an erroneous implementation $Q$ specified
as follows.
\begin{eqnarray*}
P & = & a \then P_1 \intchoice b \then P_1 \intchoice c \then P_1
\\
P_1 & = & a \then P \extchoice b\then P
\\ & &  \\
Q & = & a\then Q_1 \extchoice b\then Q_1
\\
Q_1 & = & a\then Q_2 \extchoice b\then Q_2
\\
Q_2 & = & a\then Q \intchoice b\then Q
\end{eqnarray*}
Obviously, $P$'s normalised transition graph has 2 nodes, 
while $Q$'s graph has 3.
It is easy to see (and can be checked with FDR4) that $P\lessdet_T Q$, but 
$\neg(P\lessdet_F Q)$. Furthermore, it can also be shown using the FDR4 tool that
the ``test passed condition'' 
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(k))\hide \Sigma
\]
holds for $U_F(0),\dots,U_F(4)$, but fails for $U_F(5)$. This means that the
non-conformance of $Q$ cannot be detected by any test trace of length 
less or equal to 5, but is revealed (as expected from Theorem~\ref{th:failurestest})
by a trace of length 6, because the last event offered by the test $U_F(5)$ is 
refused by $Q$.
\xbox
\end{example}
%
Generalising Example~\ref{ex:pq}, it can be shown that for any pair $p,q\in\mathbb{N}$
whose greatest common divisor is 1, there exist reference processes with $p$ states
and implementation processes with $q$ states, such that non-conformance of the 
implementation
can only be detected with a trace of length $pq$.

>>>>

The following example, however, shows that the maximal length $pq$ is really required 
when testing for refinement.
\begin{example}\label{ex:pq}
Consider the CSP reference process $P$ and an erroneous implementation $Q$ specified
as follows.
\begin{eqnarray*}
P & = & a \then P_1 \intchoice b \then P_1 \intchoice c \then P_1
\\
P_1 & = & a \then P \extchoice b\then P
\\ & &  \\
Q & = & a\then Q_1 \extchoice b\then Q_1
\\
Q_1 & = & a\then Q_2 \extchoice b\then Q_2
\\
Q_2 & = & a\then Q \intchoice b\then Q
\end{eqnarray*}
Obviously, $P$'s normalised transition graph has 2 nodes, 
while $Q$'s graph has 3.
It is easy to see (and can be checked with FDR4) that $P\lessdet_T Q$, but 
$\neg(P\lessdet_F Q)$. Furthermore, it can also be shown using the FDR4 tool that
the ``test passed condition'' 
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(k))\hide \Sigma
\]
holds for $U_F(0),\dots,U_F(4)$, but fails for $U_F(5)$. This means that the
non-conformance of $Q$ cannot be detected by any test trace of length 
less or equal to 5, but is revealed (as expected from Theorem~\ref{th:failurestest})
by a trace of length 6, because the last event offered by the test $U_F(5)$ is 
refused by $Q$.
\xbox
\end{example}
%
Generalising Example~\ref{ex:pq}, it can be shown that for any pair 
$2\le p,q \in\mathbb{N}$,  
there exist reference processes with $p$ states
and implementation processes with $q$ states, such that non-conformance of the 
implementation
can only be detected with a trace of length $pq$. This is proven in the following 
theorem.

\begin{theorem}
Let $2\le p,q \in\mathbb{N}$. Then there exists a reference process $P$ and an
implementation process $Q$ with the following properties.
\begin{enumerate}
\item $G(P)$ has $p$ states.
\item $G(Q)$ has $q$ states.
\item $P\not\lessdet_T Q$.
\item $\forall s\in\trc(Q): \#s < pq\implies s\in\trc(P)$.
\end{enumerate}
As a consequence, the upper bound $pq$ for the length of traces to be tested when checking for failures refinement cannot be reduced without losing the test suite's completeness property.
\end{theorem}
\begin{proof}
Given $2\le p,q \in\mathbb{N}$, define reference process $P$ and implementation process $Q$
over alphabet $\Sigma =\{ a,b\}$ as follows.
\begin{eqnarray*}
P & = &  P(0) 
\\
P(k) & = & \big( a \then P(k) \big) \extchoice (k < p-1) \& \big( b \then P(k+1)\big)
\\ 
Q & = & Q(0)
\\
Q(k) & = & (k < q-1) \& \big( a \then Q(k+1)    \big) \extchoice 
( k = q-1)\& \big( a\then Q(0) \extchoice b\then Q(0)  \big)
\end{eqnarray*}
Using regular expression notation, the traces of $P$ can be specified as
\[
\trc(P) = \prefs\big(  (a^*b)^{p-1}a^* \big),
\]
where $\prefs(M)$ denotes the set of all prefixes of traces in $M\subset\Sigma^*$, 
including the traces of $M$ themselves.
The traces of $Q$ can be specified by
\[
\trc(Q) = \prefs\big( (a^{q-1}(a|b))^*  \big).
\]
It is easy to see that $\trc(Q)\not\subseteq\trc(P)$; 
for example, the trace $(a^{q-1}b)^p$ is in
$\trc(Q)\setminus\trc(P)$.

Let $s \in\trc(Q)$ be any trace of length $\#s = pq-1$. Then $s$ can be represented 
by $s = (a^{q-1}(a|b))^{p-1}a^{q-1} \in \prefs\big( (a^{q-1}(a|b))^*  \big)$. 
Then $s$ is also an element of $\trc(P)$, because $(a^{q-1}(a|b))^{p-1}a^{q-1}$ 
is also contained in $\prefs\big(  (a^*b)^{p-1}a^* \big)$.
\xbox
\end{proof}
